Computable Compressed Matrices\\Suplementary Material

### 2013-09-25 manuscript.tex
Introdução e método SM analisados.

@@

Data compression is traditionally used to reduce storage resources usage and/or transmission costs\cite{salomon}. Compression techniques can be classified into lossy and lossless. Examples of lossy data compression are MP3 (audio), JPEG (image) and MPEG (video). In this paper we discuss the use of lossless compression for numerical data structures such as numerical arrays to achieve compression without losing the mathematical properties of the original data.

@@

This paper will focus solely on numerical data which for the purpose of the applications is organized as   matrices. This is a most common data structure found in computational data analysis environments. The matrices compressed according to the methodologies proposed here should be able to undergo the same mathematical operations as the original uncompressed matrices, e.g. linear algebra manipulations. This way, the cost of compression is reduced to a single event of compression and no need of decompression except when displaying the results for human reading. The idea of operating with compressed arrays is relatively new\cite{yemliha2007compiler}, and it has yet to find mainstream applications to the field of numerical computations. One application which employs a form of compression is the sparse matrix linear algebra algorithms\cite{dodson1991sparse}, in this case there is no alteration in the standard encoding of the data, but only the non-zero elements of the matrices are stored and operated upon.

Larger than RAM data structures can render traditional analytical algorithms impracticable. Currently, the technique most commonly used when dealing with large matrices for numerical computations, is memory mapping\cite{van2011numpy,big}. In memory mapping the matrix is allocated in a virtual contiguous address space which extends from memory into disk. Thus, larger than memory data structures can be manipulated as if they were in memory.  This technique has a big performance penalty due to lower access speeds of disk when compared to RAM.

In this paper we present two methods for the lossless compression of (numerical) arrays. The methods involve the encoding of the numbers as strings of bits of variable length. The methods resemble the arithmetic coding\cite{bodden2007arithmetic} algorithm, but is cheaper to compute. We describe the process of compression and decompression, and study their efficiency under different applications. We also discuss the efficiency of the compression as a function of the distribution of the elements of the matrix.

@@

To maintain mathematical equivalence with the original data for any arithmetic operations, we need to  maintain the structure of the matrix, i.e., the ability to acess any element given its row $i$ and column $j$ and also the numeric nature of its elements. In order to achieve compression we decided to exploit inefficiencies in the conventional way matrices are allocated in memory. The examples in this paper will be restricted to matrices with positive integer elements.

The compression method is as follows. Let $M_{r \times c}$ be a matrix, in which $r$ is the number of rows and $c$ the number of columns. Each element of this matrix, called $m_{ij}$, is a positive integer. In digital computers, all information is stored as binary code (base 2 numbers). However the conventional way to store arrays of integers is on a memory block sequence of fixed size (power of 2 numbers of bit), one for each element. The maximum size of a block is equal to the word size of the processor, which for most current CPUs is 64 bits. Some special number such as complex number may be encoded as two blocks instead of one. The size of the chunk of memory allocated to each number will determine their maximum  size (for integers) or their precision (for floating-point numbes). So for matrix $M$, the total memory allocated, assuming chunks of 64 bits, is given by $\mathcal{B} = r \times c \times 64$.

@@

\begin{equation} \label{eq:01}
 b(\sup M) \approx \begin{cases}
	1, &  \text{ if } \sup M \in \{0,1\} \\ 
	\lfloor \log_2(\sup M)  \rfloor + 1,  & \text{ if } \sup M > 1  
	\end{cases}
\end{equation}

The allocation of memory still happens in the usual way, i.e., in fixed size 64-bit chunks, only that now, in the space required for a single 64 bit integer, we can store for example, an entire $8\times 8$ matrix of $0_{10}$ and $1_{10}$.

@@

\begin{table}[h]
 \centering
 \caption{Some elements of $M$ represented in binary base.}
 \begin{tabular}{cccc} 
  \hline 
  Element & Value  & Binary & Bit length\\
  \hline
  $M_{1,1}$ & 900  & 1110000100 & 10\\
  $M_{1,2}$ & 1023 & 1111111111 & 10\\
  $M_{1,3}$ & 721  & 1011010001 & 10\\
  $M_{1,4}$ & 256  & 100000000  & 9\\
  $M_{1,5}$ & 1    & 1          & 1\\
  $M_{1,6}$ & 10   & 1010       & 4\\
  $M_{1,7}$ & 700  & 1010111100 & 10\\
  $M_{1,8}$ & 20   & 10100      & 5\\
  \hline
 \end{tabular}
 \label{tab:01}
\end{table}

@@

According to the SM method, having determined that each element will require at most 10 bits, we can divide the memory block corresponding to a single 64 bit integer into six 10-bit chunks which can each hold a single element of $M$. These 64-bit blocks will be called a bitstring. The remaining 4 bits will be used later. The number of bitstrings needed will be $\lfloor \frac{\dim(M) *b(\sup M)} {64} \rfloor +1$, where $dim(M)$ is the dimension of the matrix or its number of elements.

@@

\begin{enumerate}
 \item Element $M_{1,1}=900 = m_{1,1} = 1110000100$ is stored in the first 10-bit chunk of the element strip $bitstring[1]$ , which corresponds to bits 0 to 9 (read from right to left).
\begin{equation*}\label{eq:04}
bitstring_1 = 000000000000000000000000000000000000000000000000000000\underbrace{1110000100}_{900} 
\end{equation*}
 \item Element $M_{1,2}=1023$ is allocated in the second chunk, from bit 10 to bit 19.
\begin{equation*}\label{eq:05}
 bitstring_1 = 00000000000000000000000000000000000000000000\underbrace{1111111111}_{1023}\underbrace{1110000100}_{900} 
\end{equation*}
 \item Repeat for elements $M_{1,i}$ with $i=1,\ldots,6$ which are stored on the remaining chunks.
 \begin{equation*}\label{eq:06}
bitstring_1 = 0000\underbrace{0000001010}_{10}\underbrace{0000000001}_{1}\underbrace{0100000000}_{256}\underbrace{1011010001}_{721}\underbrace{1111111111}_{1023}\underbrace{1110000100}_{900} 
 \end{equation*}
 \item Element $M_{1,7}=700=1010111100$ does not fit on the remaining 4 bits of the first bitstring. So it will straddle two bitstrings, i.e., it is divided in two segments $a$ and $b$, $a$ is written on the first bitstring and $b$ on the second.
 \begin{equation*}\label{eq:07}
  \underbrace{\underbrace{0000010100}_{20}\overbrace{101011}^{b}}_{bitstring_2}|\underbrace{\overbrace{1100}^{a}\underbrace{0000001010}_{10}\underbrace{0000000001}_{1}\hdots}_{bitstring_1}
\end{equation*}
\end{enumerate}

 Please notice that bitstrings are written from right to left.

@@

Thus the compressed matrix $m = M_2$ requires less memory than the conventional storage of $M$ as a 64-bit integer array.

@@

%\section*{Figure Legends}
%\begin{figure}[!ht]
%\begin{center}
%%\includegraphics[width=4in]{figure_name.2.eps}
%\end{center}
%\caption{
%{\bf Bold the first sentence.}  Rest of figure 2  caption.  Caption 
%should be left justified, as specified by the options to the caption 
%package.
%}
%\label{Figure_label}
%\end{figure}

>>>>

\usepackage{todonotes}

@@

Data compression is traditionally used to reduce storage resources usage and/or 
transmission costs\cite{salomon}. Compression techniques can be classified into 
lossy and lossless. Examples of lossy data compression are MP3 (audio), JPEG 
(image) and MPEG (video). In this paper, we discuss the use of lossless 
compression for numerical data structures such as numerical arrays to achieve 
compression without losing the mathematical properties of the original data.

@@

This paper will focus solely on numerical data which for the purpose of the 
applications is organized as matrices. This is a most common data structure 
found in computational data analysis environments. The matrices compressed 
according to the methodologies proposed here should be able to undergo the same 
mathematical operations as the original uncompressed matrices, e.g. linear 
algebra manipulations. This way, the cost of compression is reduced to a single 
event of compression and no need of decompression, except when displaying the 
results for human reading. The idea of operating with compressed arrays is 
relatively new\cite{yemliha2007compiler}, and it has yet to find mainstream 
applications to the field of numerical computations. One application which 
employs a form of compression is the sparse matrix linear algebra 
algorithms\cite{dodson1991sparse}, in this case there is no alteration in the 
standard encoding of the data, but only the non-zero elements of the matrices 
are stored and operated upon.

Larger than RAM data structures can render traditional analytical algorithms 
impracticable. Currently, the technique most commonly used when dealing with 
large matrices for numerical computations, is memory 
mapping\cite{van2011numpy,big}. In memory, mapping the matrix is allocated in a 
virtual contiguous address space which extends from memory into disk. Thus, 
larger than memory data structures can be manipulated as if they were in memory. 
This technique has a big performance penalty due to lower access speeds of disk 
when compared to RAM.

In this paper, we present two methods for the lossless compression of 
(numerical) arrays. The methods involve the encoding of the numbers as strings 
of bits of variable length. The methods resemble the arithmetic 
coding\cite{bodden2007arithmetic} algorithm, but is cheaper to compute. We 
describe the process of compression and decompression, and study their 
efficiency under different applications. We also discuss the efficiency of the 
compression as a function of the distribution of the elements of the matrix. 
\todo{Inseri isso} Also presented are some results in relation to the 
traditional method of matrix allocation. We also consider the 
arithmetic operations to measure the efficiency of the proposed methodology.

@@

To maintain mathematical equivalence with the original data for any arithmetic 
operations, we need to maintain the structure of the matrix, ie, the ability 
to access any element given its row $i$ and column $j$ and also the numeric 
nature of its elements. In order to achieve compression, we decided to exploit 
inefficiencies in the conventional way matrices are allocated in memory. 
\todo{adicionei aqui} The analyzes presented in this article, in the first 
moment, are related to arrays of positive integers, but can be applied to 
integers and real numbers with some adaptations, which are presented in due 
course. \todo{removi isso The examples in this paper will be restricted to 
matrices with positive integer elements, but}

The compression method is as follows. Let $M_{r \times c}$ be a matrix, in which 
$r$ is the number of rows and $c$ the number of columns. Each element of this 
matrix, called $m_{ij}$, is a positive integer. In digital computers, all 
information is stored as binary code (base 2 numbers). However, the conventional 
way to store arrays of integers is on a memory block sequence of fixed size 
(power of 2 numbers of bit), one for each element. The maximum size of a block 
is equal to the word size of the processor, which for most current CPUs is 64 
bits. Some special number such as complex number may be encoded as two blocks 
instead of one. The size of the chunk of memory allocated to each number will 
determine their maximum  size (for integers) or their precision (for 
floating-point numbers). So for matrix $M$, the total memory allocated, 
assuming chunks of 64 bits, is given by $\mathcal{B} = r \times c \times 64$.

@@

\begin{equation} \label{eq:01}
 b(\sup M) \approx \begin{cases}
	1, &  \text{ if } \sup M \in \{0,1\} \\ 
	\lfloor \log_2(\sup M)  \rfloor + 1,  & \text{ if } \sup M > 1  
	\end{cases}
\end{equation}\todo{Será que não devemos trocar $\approx$ por = aqui? Note que 
o número de bits não é aproximado e sim o exato.}

The allocation of memory still happens in the usual way, ie, in fixed size 
64-bit chunks, only that now, in the space required for a single 64 bit integer, 
we can store for example, an entire $8\times 8$ matrix of $0_{10}$ and 
$1_{10}$.

@@

\begin{table}[h]
 \centering
 \caption{The first 8 elements of $M$ represented in binary base.}
 \begin{tabular}{cccc} 
  \hline 
  Element & Value  & Binary & Bit length\\
  \hline
  $M_{1,1}$ & 900  & 1110000100 & 10\\
  $M_{1,2}$ & 1023 & 1111111111 & 10\\
  $M_{1,3}$ & 721  & 1011010001 & 10\\
  $M_{1,4}$ & 256  & 100000000  & 9\\
  $M_{1,5}$ & 1    & 1          & 1\\
  $M_{1,6}$ & 10   & 1010       & 4\\
  $M_{1,7}$ & 700  & 1010111100 & 10\\
  $M_{1,8}$ & 20   & 10100      & 5\\
  \hline
 \end{tabular}
 \label{tab:01}
\end{table}

@@

According to the SM method, having determined that each element will require at 
most 10 bits, we can divide the memory block corresponding to a single 64 bit 
integer into six 10-bit chunks, which can each hold a single element of $M$. 
These 64-bit blocks will be called a bitstring. The remaining 4 bits will be 
used later. The number of bitstrings needed will be $\lfloor \frac{\dim(M) 
*b(\sup M)} {64} \rfloor +1$, where $dim(M)$ is the dimension of the matrix or 
its number of elements.

@@

\begin{enumerate}
 \item Element $M_{1,1}=900 = m_{1,1} = 1110000100$ is stored in the first 10-bit chunk of the element strip $bitstring[1]$ , which corresponds to bits 0 to 9 (read from right to left).
\begin{equation*}\label{eq:04}
bitstring_1 = 000000000000000000000000000000000000000000000000000000\underbrace{1110000100}_{900} 
\end{equation*}
 \item Element $M_{1,2}=1023$ is allocated in the second chunk, from bit 10 to bit 19.
\begin{equation*}\label{eq:05}
 bitstring_1 = 00000000000000000000000000000000000000000000\underbrace{1111111111}_{1023}\underbrace{1110000100}_{900} 
\end{equation*}
 \item Repeat for elements $M_{1,i}$ with $i=1,\ldots,6$ which are stored on the remaining chunks.
 \begin{equation*}\label{eq:06}
bitstring_1 = 0000\underbrace{0000001010}_{10}\underbrace{0000000001}_{1}\underbrace{0100000000}_{256}\underbrace{1011010001}_{721}\underbrace{1111111111}_{1023}\underbrace{1110000100}_{900} 
 \end{equation*}
 \item Element $M_{1,7}=700=1010111100$ does not fit on the remaining 4 bits of 
the first bitstring. So it will straddle two bitstrings, ie, it is divided in 
two segments $a$ and $b$, $a$ is written on the first bitstring and $b$ on the 
second.
 \begin{equation*}\label{eq:07}
  \underbrace{\underbrace{0000010100}_{20}\overbrace{101011}^{b}}_{bitstring_2}|\underbrace{\overbrace{1100}^{a}\underbrace{0000001010}_{10}\underbrace{0000000001}_{1}\hdots}_{bitstring_1}
\end{equation*}
\end{enumerate}

 Please, notice that bitstrings are written from right to left.

@@

Thus the compressed matrix $m = M_2$ requires less memory than the conventional 
storage of $M$ as a 64-bit integer array. An illustration of the compression 
process and the allocation of the elements of matrix M in the matrix m is shown 
in Figure \ref{fig:01}. Note that the matrix m has fewer columns that the 
matrix M, then it demands less memory to be represented. We also emphasize that 
the bits which represent the value 700 is divided in two parts and stored in 
different strips in the matrix m.

@@

\section*{Figure Legends}

\begin{figure}[htp]
  \includegraphics[scale=0.3,clip]{fig01.pdf}
  \caption{Representation of the compression process and allocation of 
elements of matrix M in the matrix m. The strip of value 700 is divided in two 
parts and stores in different strips in the matrix m.}
  \label{fig:01}
\end{figure}

%\begin{figure}[!ht]
%\begin{center}
%%\includegraphics[width=4in]{figure_name.2.eps}
%\end{center}
%\caption{
%{\bf Bold the first sentence.}  Rest of figure 2  caption.  Caption 
%should be left justified, as specified by the options to the caption 
%package.
%}
%\label{Figure_label}
%\end{figure}
